Generating Random Logic Programs Using Constraint Programming

### 2019-12-28 paper/paper.tex
References + formatting

\documentclass{article}
\usepackage[UKenglish]{babel}
\usepackage[UKenglish]{isodate}
\usepackage[utf8]{inputenc}
\usepackage{fullpage}
\usepackage[ruled,vlined]{algorithm2e}
\usepackage{amsmath}
\usepackage{amsthm}
\usepackage{amssymb}

\newtheorem{constraint}{Constraint}
\theoremstyle{definition}
\newtheorem{definition}{Definition}

\newcommand{\variable}[1]{\texttt{\textup{#1}}}
\newcommand{\predicates}{\mathcal{P}}
\newcommand{\variables}{\mathcal{V}}
\newcommand{\constants}{\mathcal{C}}
\newcommand{\tokens}{\mathcal{T}}
\newcommand{\arities}{\mathcal{A}}
\newcommand{\maxArity}{\textsc{max\_arity}}
\newcommand{\maxNumNodes}{\textsc{max\_num\_nodes}}
\newcommand{\maxNumClauses}{\textsc{max\_num\_clauses}}

\author{Paulius Dilkas}
\title{Generating Random Logic Programs Using Constraint Programming}

\begin{document}
\maketitle

@@

We will often use $\Box$ as a special domain value to indicate some kind of
exception.

@@

\begin{itemize}
\item Make negative cycle detection use the graph representation.
\item Show that the set of all ProbLog programs is equal to the set of programs
  I can generate (alternatively, show that, given any ProbLog program, there are
  parameter values high enough to generate it).
\item Given fixed parameters, use combinatorial arguments to calculate how many
  different programs there are and check that I'm generating the same number.
\item Formal definition (here and in the predicate invention paper): two
  predicates are independent if all of their groundings are independent.
\item Describe: entailment checking for cycles and negative cycles.
\item A constraint for logical equivalence.
\item Rename everything into something more appropriate.
\end{itemize}

Both determined $\implies$ fail(). Both determined but at least is one masked by
a (probable/determined) mask $\implies$ nothing. One determined $\implies$ the
other one cannot exist.

@@

Parameters:
\begin{itemize}
\item maximum number of solutions
\item \maxNumNodes{} (in the tree representation of a clause)
\item \maxNumClauses{}
\item option to forbid all cycles or just negative cycles
\item a list of probabilities that are randomly assigned to clauses: $\{ 0.1,
  0.2, \dots, 0.9, 1, 1, 1, 1, 1, 1 \}$.
\item a list of predicates $\predicates{}$,
\item a list of their arities $\arities{}$,
  \begin{itemize}
  \item $\maxArity{} = \max \arities{}$.
  \end{itemize}
\item a list of variables $\variables{}$,
\item and a list of constants $\constants{}$.
\end{itemize}

We also define $\tokens{} = \{ \neg, \land, \lor, \top \}$.

Decision variables:
% TODO: specify the decision variables inside clauses and their heads
\begin{itemize}
\item \variable{Body[] bodiesOfClauses}
\item \variable{Head[] headsOfClauses}: a list of length \maxNumClauses{}
\end{itemize}

@@

\begin{constraint}[Each predicate gets at least one clause]
  Let $P = \{ h.\variable{predicate} \mid h \in \variable{clauseHeads} \}$. Then
  \[
    \variable{nValues}(P) =
    \begin{cases}
      \variable{numPredicates} + 1 & \text{if } \variable{count}(\Box, P) > 0 \\
      \variable{numPredicates} & \text{otherwise.}
    \end{cases}
  \]
\end{constraint}

\begin{constraint}
  Let $\prec$ be any total order defined over bodies of clauses, and let
  $\preceq$ be its extension with equality (in the same way as $\le$ extends
  $<$).
  \[
    \variable{headsOfClauses}[i-1].\variable{predicate} =
    \variable{headsOfClauses}[i].\variable{predicate} \implies
    \variable{bodiesOfClauses}[i-1] \preceq \variable{bodiesOfClauses}[i].
  \]
\end{constraint}
For example, $\preceq$ can be implemented as \variable{lexLessEq} over the
decision variables of each body.
% TODO: might need to include clause heads as well.

@@

\begin{definition}
  An \emph{atom} is a $\variable{predicate} \in \tokens{} \cup \predicates{}$ and a
  list of \variable{arguments} of length \maxArity{} in $\variables{} \cup
  \constants{} \cup \{ \Box \}$, where $\Box$ means the position is either
  reserved for a variable, or disabled. The atom's \variable{arity} is a number
  in $[0, \maxArity{}]$ defined by a \variable{table} constraint,
  according to the \variable{predicate}.
\end{definition}

@@

\begin{definition}
  The body of a clause is defined by:
  \begin{itemize}
  \item \variable{treeStructure}: list of length \maxNumNodes{} with
    domain $[0, \maxNumNodes{}]$.
    \begin{itemize}
    \item $\variable{treeStructure}[i] = i$: the $i$-th node is a root.
    \item $\variable{treeStructure}[i] = j$: the $i$-th node's parent is node $j$.
    \end{itemize}
  \item \variable{treeValues}: \maxNumNodes{} atoms.
  \end{itemize}
\end{definition}
% TODO: examples

@@

\begin{constraint}
  \variable{treeStructure} represents \variable{numTrees} trees.
\end{constraint}

@@

\begin{constraint}
  \variable{treeStructure} is sorted.
\end{constraint}

@@

@@

\begin{constraint}
  For each variable $v \in \variable{variables}$, $v \le \variable{arity}$.
\end{constraint}

@@

\begin{constraint}[Connecting the two lists]
  $\sum_{v \in \variable{variables}} v = \variable{count}(\Box,
  \variable{constants}) + \variable{arity} - \maxArity{}$.
\end{constraint}

@@

@@

@@

>>>>

\documentclass[runningheads]{llncs}
\usepackage{graphicx}
%\usepackage[UKenglish]{babel}
%\usepackage[UKenglish]{isodate}
%\usepackage[utf8]{inputenc}
\usepackage[ruled,vlined]{algorithm2e}
\usepackage{amssymb}
\usepackage{mathtools}
\usepackage[capitalise]{cleveref}

\newtheorem{constraint}{Constraint}

\newcommand{\variable}[1]{\texttt{\textup{#1}}}
\newcommand{\predicates}{\mathcal{P}}
\newcommand{\variables}{\mathcal{V}}
\newcommand{\constants}{\mathcal{C}}
\newcommand{\tokens}{\mathcal{T}}
\newcommand{\arities}{\mathcal{A}}
\newcommand{\maxArity}{\mathcal{M}_{\mathcal{A}}}
\newcommand{\maxNumNodes}{\mathcal{M}_{\mathcal{N}}}
\newcommand{\maxNumClauses}{\mathcal{M}_{\mathcal{C}}}

\begin{document}

\title{Generating Random Logic Programs Using Constraint Programming}
\author{Paulius Dilkas\orcidID{0000-1111-2222-3333}}
\authorrunning{P. Dilkas}
% First names are abbreviated in the running head.
% If there are more than two authors, 'et al.' is used.

\institute{University of Edinburgh, Edinburgh, UK\\
\email{p.dilkas@sms.ed.ac.uk}}
\maketitle

\begin{abstract}
The abstract should briefly summarize the contents of the paper in
150--250 words.

\keywords{Constraint Programming \and Logic Programming \and Probabilistic Logic
  Programming.}
\end{abstract}

@@

We will often use $\Box$ as a special domain value to indicate some kind of
exception. We also use Choco \cite{choco}. This works with both Prolog
\cite{DBLP:books/daglib/0041598} and ProbLog \cite{DBLP:conf/ijcai/RaedtKT07}.
Tested with SWI-Prolog \cite{DBLP:journals/tplp/WielemakerSTL12}.

Inference options to explore. Logspace vs normal space. Symbolic vs
non-symbolic. Propagate evidence (might be irrelevant)? Propagate weights?
Supported knowledge compilation techniques: sdd, sddx, bdd, nnf, ddnnf, kbest,
fsdd, fbdd.

@@

\begin{itemize}
\item Given fixed parameters, use combinatorial arguments to calculate how many
  different programs there are and check that I'm generating the same number.
\item Formal definition (here and in the predicate invention paper): two
  predicates are independent if all of their groundings are independent.
\item A constraint for logical equivalence.
\item Show that the set of all ProbLog programs is equal to the set of programs
  I can generate (alternatively, show that, given any ProbLog program, there are
  parameter values high enough to generate it).
\item Perhaps negative cycle detection could use the same graph as the
  independence propagator? If we extend each domain to {-1, 0, 1}, but that
  might make propagation weaker or slower.
\end{itemize}

@@

Parameters:
\begin{itemize}
\item maximum number of solutions,
\item $\maxNumNodes{}$: maximum number of nodes in the tree representation of a
  clause,
\item $\maxNumClauses{}$: maximum number of clauses in a program,
\item option to forbid all cycles or just negative cycles,
\item a list of probabilities that are randomly assigned to clauses,
\item a list of predicates $\predicates{}$,
\item a list of their arities $\arities{}$,
  \begin{itemize}
  \item maximum arity $\maxArity{} \coloneqq \max \arities{}$.
  \end{itemize}
\item a list of variables $\variables{}$,
\item and a list of constants $\constants{}$.
\end{itemize}

We also define $\tokens{} = \{ \neg, \land, \lor, \top \}$. All decision
variables of the model are contained in two arrays of length $\maxNumClauses{}$:
\begin{itemize}
\item \variable{Body[] bodiesOfClauses}
\item \variable{Head[] headsOfClauses}
\end{itemize}

@@

\begin{constraint}
  For $i = 0, \dots, \maxNumClauses - 1$, let $p_i =
  \variable{headsOfClauses}[i].\variable{predicate}$. Then
  \[
    \variable{sort}((p_i)_{i=0}^{\maxNumClauses - 1}, (p_i)_{i=0}^{\maxNumClauses
      - 1}),
  \]
  i.e., all clauses describing the same predicate are listed together and
  sorted.
\end{constraint}

\begin{constraint}
  Each predicate gets at least one clause. Let $P = \{ h.\variable{predicate}
  \mid h \in \variable{clauseHeads} \}$. Then
  \[
    \variable{nValues}(P) =
    \begin{cases}
      \variable{numPredicates} + 1 & \text{if } \variable{count}(\Box, P) > 0 \\
      \variable{numPredicates} & \text{otherwise.}
    \end{cases}
  \]
\end{constraint}

\begin{constraint}
  Let $\prec$ be any total order defined over bodies of clauses, and let
  $\preceq$ be its extension with equality (in the same way as $\le$ extends
  $<$). If
  \[
    \variable{headsOfClauses}[i-1].\variable{predicate} =
    \variable{headsOfClauses}[i].\variable{predicate},
  \]
  then $\variable{bodiesOfClauses}[i-1] \preceq \variable{bodiesOfClauses}[i]$.
\end{constraint}
For example, $\preceq$ can be implemented as \variable{lexLessEq} over the
decision variables of each body.
% TODO: might need to include clause heads as well.

@@

\begin{definition} \label{def:atom}
  An \emph{atom} is a $\variable{predicate} \in \tokens{} \cup \predicates{}$ and a
  list of \variable{arguments} of length $\maxArity{}$ in $\variables{} \cup
  \constants{} \cup \{ \Box \}$, where $\Box$ means the position is either
  reserved for a variable, or disabled. The atom's \variable{arity} is a number
  in $[0, \maxArity{}]$ defined by a \variable{table} constraint,
  according to the \variable{predicate}.
\end{definition}

@@

\begin{definition}
  The body of a clause is defined by:
  \begin{itemize}
  \item \variable{treeStructure}: list of length $\maxNumNodes{}$ with
    domain $[0, \maxNumNodes{} - 1]$.
    \begin{itemize}
    \item $\variable{treeStructure}[i] = i$: the $i$-th node is a root.
    \item $\variable{treeStructure}[i] = j$: the $i$-th node's parent is node $j$.
    \end{itemize}
  \item \variable{treeValues}: $\maxNumNodes{}$ atoms.
  \end{itemize}
\end{definition}
% TODO: examples

@@

\begin{constraint}
  $\variable{tree}(\variable{treeStructure}, \variable{numTrees})$, i.e.,
  \variable{treeStructure} represents \variable{numTrees} trees (dominator-based
  filtering \cite{DBLP:conf/cp/FagesL11}).
\end{constraint}

@@

\begin{constraint}
  $\variable{sort}(\variable{treeStructure}, \variable{treeStructure})$, i.e.,
  \variable{treeStructure} is sorted.
\end{constraint}

@@

Our definition of a head of a clause is more restrictive than \cref{def:atom}.

@@

\begin{constraint}
  For each $v \in \variable{variables}$, $v \le \variable{arity}$.
\end{constraint}

@@

\begin{constraint}
  Connecting the two lists:
  \[
    \sum_{v \in \variable{variables}} v = \variable{count}(\Box,
    \variable{constants}) + \variable{arity} - \maxArity{}.
  \]
\end{constraint}

@@

\section{Entailment Checking for Negative/All Cycles}

\begin{enumerate}
\item Let $C$ be a set of clauses such that their bodies and predicates in their
  heads are fully determined.
\item If $C = \emptyset$, return UNDEFINED.
\item Construct an adjacency list representation of a graph where vertices
  represent predicates. Each edge is either \emph{positive} or \emph{negative}.
  There is an edge from $p$ to $q$ if $q$ appears in the body of a predicate
  with $p$ as its head. The edge is negative if, when traversing the tree to
  reach some instance of $q$, we pass through a $\neg$ node. Otherwise, it's
  positive.
\item Run a modified cycle detection algorithm that detects all cycles that have
  at least one negative edge.
\item If we found a cycle, return FALSE.
\item If $C$ encompasses all clauses, return TRUE.
\item Return UNDEFINED.
\end{enumerate}

@@

% Both determined $\implies$ fail(). Both determined but at least is one masked by
% a (probable/determined) mask $\implies$ nothing. One determined $\implies$ the
% other one cannot exist.

@@

\section*{Acknowledgments}

The author would like to thank Vaishak Belle for his comments. This work was
supported by the EPSRC Centre for Doctoral Training in Robotics and Autonomous
Systems, funded by the UK Engineering and Physical Sciences Research Council
(grant EP/S023208/1).

\bibliographystyle{splncs04}
\bibliography{paper}
